---
title: "Reconcile Methylation Array Data Sample Info"
subtitle: "Mapping samples between files and getting demographic info"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
mainfont: "DejaVu Sans"
fontsize: 10pt
header-includes:
  - \usepackage{titlesec}
  - \usepackage{xcolor}
  - \definecolor{darkblue}{HTML}{00008B}
  - \titleformat{\section}{\color{darkblue}\Large\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\color{darkblue}\large\bfseries}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\color{darkblue}\normalsize\bfseries}{\thesubsubsection}{1em}{}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 4,
  fig.align  = "center"
)
# global plot theme so all ggplots inherit readable font sizes
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```
```{r libraries, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(ggVennDiagram)
source("/home/ec2-user/AMP-AD_methylation_harmonization/R/harmonize_metadata.R")
source("/home/ec2-user/AMP-AD_methylation_harmonization/R/build_idat_manifest.R")
```
```{r demographics-helper, include=FALSE}
# Summarize the core harmonized demographic columns for one cohort.
# `df` is expected to be a per-specimen (methylation) data frame. Demographic
# columns are collapsed to one row per individualID before tabulating so
# individuals with multiple specimens aren't counted multiple times; tissue
# is left at the specimen level so it reflects the real sample count.
summarize_demographics <- function(df, cohort_name,
                                    vars = c("sex","race","isHispanic","apoe4Status","Braak", "tissue"),
                                    var_labels = c("Sex","Race","Hispanic Ethnicity","apoe4 Status","Braak Stage", "Specimen Tissue"),
                                    id_col = "individualID") {
  keep <- vars %in% names(df)
  vars <- vars[keep]
  var_labels <- var_labels[keep]

  df_by_individual <- df[!duplicated(df[[id_col]]), ]

  for (v in seq_along(vars)) {
    v_name <- var_labels[v]
    cat("\n", v_name, "\n\n")
    tbl_df <- if (vars[v] == "tissue") df else df_by_individual
    tbl <- as.data.frame(table(tbl_df[[vars[v]]], useNA = "ifany"))
    names(tbl) <- c(v_name, "Count")
    print(kable(tbl))
    cat("\n")
  }

  if ("ageDeath" %in% names(df)) {
    age_numeric <- suppressWarnings(as.numeric(ifelse(df_by_individual$ageDeath == "90+", "90", df_by_individual$ageDeath)))
    cat("\n Age at Death (numeric summary; '90+' top-coded to 90)\n\n")
    age_summary <- summary(age_numeric)
    age_tbl <- data.frame(Statistic = names(age_summary), Value = as.numeric(age_summary))
    print(kable(age_tbl))
    cat("\n")
  }
}

# Venn diagram of metadata/IDAT overlap for one cohort.
plot_overlap_venn <- function(metadata_samples, idat_samples, cohort_name) {
  venn_list <- list("Metadata" = metadata_samples, "IDAT" = idat_samples)

  ggVennDiagram(venn_list, label = "both", label_alpha = 0) +
    scale_fill_gradient(low = "#eaf2fc", high = "#2a78d6") +
    scale_x_continuous(expand = expansion(mult = 0.2)) +
    labs(title = paste0(cohort_name, ": Metadata vs. IDAT File Overlap")) +
    theme_void(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none", plot.margin = unit(c(0.5, 15, 1, 5), "cm")) +
    coord_cartesian(clip = "off") 
}
```
# MSBB
| Synapse ID | File name |
|---|---|
| syn21447661 | ROU_13169_B01_CUS_MethylEPIC.illumina.beta.txt |
| syn52073071 | Raw data folder |
| syn22157601 | MSBB_assay_MethylationArray_metadata.csv |
| syn21893059 | MSBB_biospecimen_metadata.csv |
| syn6101474 | MSBB_individual_metadata.csv |

## Raw Data availability 
```{r MSBB, echo = FALSE, results = 'hide'}
msbb_dir <- '/home/ec2-user/data/methyl_harmonization/methyl_data/MSBB'
msbb_out_dir <- '/home/ec2-user/data/methyl_harmonization/MSBB/metadata/'
dir.create(msbb_out_dir, recursive = T)
msbb_meth_df <- read.csv(file.path(msbb_dir,'metadata/MSBB_assay_MethylationArray_metadata.csv'))
msbb_bio_df <- read.csv(file.path(msbb_dir,'metadata/MSBB_biospecimen_metadata.csv'))
#length(intersect(msbb_meth_df$specimenID, msbb_bio_df$specimenID)) #All 201 samples accounted for
msbb_ind_df <- read.csv(file.path(msbb_dir,'metadata/MSBB_individual_metadata_harmonized.csv'))
z <- msbb_bio_df[msbb_bio_df$specimenID %in% msbb_meth_df$specimenID & msbb_bio_df$exclude != TRUE,] #No samples flagged to be excluded 


#map the specimenID to individualID and add covariate information 
msbb_bio_df_subset <- msbb_bio_df[msbb_bio_df$specimenID %in% msbb_meth_df$specimenID,]  
msbb_meth_df$individualID <- msbb_bio_df_subset$individualID[match(msbb_meth_df$specimenID, msbb_bio_df_subset$specimenID)]
msbb_ind_df <- msbb_ind_df[msbb_ind_df$individualID %in% msbb_meth_df$individualID, ]


#Add tissue info to methylation dataset
msbb_meth_df$tissue <- msbb_bio_df_subset$tissue[match(msbb_meth_df$specimenID, msbb_bio_df_subset$specimenID)]

#Add tissue info to individual metadata
msbb_ind_df$tissue <- msbb_bio_df_subset$tissue[match(msbb_ind_df$individualID, msbb_bio_df_subset$individualID)]

# Validate harmonization
validate_harmonized(msbb_ind_df,"MSBB")

# Check if green and red .idat files present for all samples
msbb_grn_file_names <- list.files(file.path(msbb_dir, 'raw_data'), pattern = '_Grn\\.idat$', recursive = TRUE)
msbb_red_file_names <- list.files(file.path(msbb_dir, 'raw_data'), pattern = '_Red\\.idat$', recursive = TRUE)

msbb_grn_sample_names <- sub('_Grn\\.idat$', '', basename(msbb_grn_file_names))
msbb_red_sample_names <- sub('_Red\\.idat$', '', basename(msbb_red_file_names))

msbb_sample_names <- paste0(msbb_meth_df$Sentrix_ID, '_', msbb_meth_df$Sentrix_Row_Column)

length(intersect(msbb_sample_names, msbb_grn_sample_names))
length(intersect(msbb_sample_names, msbb_red_sample_names))

length(unique(msbb_ind_df$individualID))

# Samples with metadata but missing a Grn and/or Red IDAT file
msbb_missing_idat <- union(
  setdiff(msbb_sample_names, msbb_grn_sample_names),
  setdiff(msbb_sample_names, msbb_red_sample_names)
)

# IDAT files present on disk with no matching metadata row
msbb_missing_metadata <- union(
  setdiff(msbb_grn_sample_names, msbb_sample_names),
  setdiff(msbb_red_sample_names, msbb_sample_names)
)

cat(length(msbb_missing_idat), "sample(s) in metadata with no matching IDAT file(s):\n")
print(msbb_missing_idat)

cat(length(msbb_missing_metadata), "IDAT sample(s) on disk with no matching metadata row:\n")
print(msbb_missing_metadata)

msbb_idat_sample_names <- union(msbb_grn_sample_names, msbb_red_sample_names)
plot_overlap_venn(msbb_sample_names, msbb_idat_sample_names, "MSBB")

# Remove samples missing IDAT files before writing out and summarizing
msbb_meth_df <- msbb_meth_df[!msbb_sample_names %in% msbb_missing_idat, ]
msbb_ind_df <- msbb_ind_df[msbb_ind_df$individualID %in% msbb_meth_df$individualID, ]

# Write out processed data
write.table(msbb_ind_df, file = file.path(msbb_out_dir, 'MSBB_individual_metadata_processed.csv'), sep = ',', col.names = T, row.names = F)
write.table(msbb_meth_df, file = file.path(msbb_out_dir, 'MSBB_assay_methylationArray_metadata_processed.csv'), sep = ',', col.names = T, row.names = F)

# Manifest: specimenID/individualID <-> green/red IDAT file names
msbb_manifest <- build_idat_manifest(msbb_meth_df, "Sentrix_ID", "Sentrix_Row_Column",
                                      msbb_grn_file_names, msbb_red_file_names)
write.table(msbb_manifest, file = file.path(msbb_out_dir, 'MSBB_idat_manifest.csv'), sep = ',', col.names = T, row.names = F)
```
`r length(intersect(msbb_sample_names, msbb_grn_sample_names))` of `r length(msbb_sample_names)` Green IDAT files and `r length(intersect(msbb_sample_names, msbb_red_sample_names))` of `r length(msbb_sample_names)` Red IDAT files present for MSBB cohort.

`r length(msbb_missing_idat)` sample(s) in the metadata have no matching IDAT file `r if (length(msbb_missing_idat)) paste0(':', " ", paste(msbb_missing_idat, collapse = ", ")) else ""`.

`r length(msbb_missing_metadata)` IDAT sample(s) on disk have no matching metadata row: `r if (length(msbb_missing_metadata)) paste0(':', " ", paste(msbb_missing_metadata, collapse = ", ")) else ""`.

## Demographics

```{r MSBB-demographics, echo = FALSE, results = "asis"}
demo_vars <- c("sex","race","isHispanic","apoe4Status","Braak","ageDeath")
demo_vars <- demo_vars[demo_vars %in% names(msbb_ind_df)]
msbb_meth_df[demo_vars] <- msbb_ind_df[match(msbb_meth_df$individualID, msbb_ind_df$individualID), demo_vars]

summarize_demographics(msbb_meth_df, "MSBB")
```

# ROSMAP
| Synapse ID | File name |
|---|---|---|
| syn3168763 | ROSMAP_arrayMethylation_imputed.tsv.gz | 
| syn5850422 | ROSMAP_arrayMethylation_raw.gz | 
| syn3168775 | ROSMAP_arrayMethylation_metaData.tsv | 
| syn23569437 | ROSMAP_assay_methylationArray_metadata.csv | 
| syn21323366 | ROSMAP_biospecimen_metadata.csv | 
| syn3191087 | ROSMAP_clinical.csv | 

## Raw Data availability 
```{r ROSMAP, echo = FALSE, results = 'hide'}
rosmap_dir <- '/home/ec2-user/data/methyl_harmonization/methyl_data/ROSMAP'
rosmap_out_dir <- '/home/ec2-user/data/methyl_harmonization/ROSMAP/metadata/'
dir.create(rosmap_out_dir, recursive = T)
rosmap_meth_df <- read.csv(file.path(rosmap_dir,'metadata/ROSMAP_assay_methylationArray_metadata.csv'))
rosmap_bio_df <- read.csv(file.path(rosmap_dir,'metadata/ROSMAP_biospecimen_metadata.csv'))
#length(intersect(rosmap_meth_df$specimenID, rosmap_bio_df$specimenID)) #All 748 samples accounted for
rosmap_ind_df <- read.csv(file.path(rosmap_dir,'metadata/ROSMAP_clinical_harmonized.csv'))
x <- rosmap_bio_df[rosmap_bio_df$specimenID %in% rosmap_meth_df$specimenID & is.na(rosmap_bio_df$exclude),] #All samples have NA as exlude 

#map the specimenID to individualID and add covariate information 
rosmap_bio_df_subset <- rosmap_bio_df[rosmap_bio_df$specimenID %in% rosmap_meth_df$specimenID,]  
rosmap_meth_df$individualID <- rosmap_bio_df_subset$individualID[match(rosmap_meth_df$specimenID, rosmap_bio_df_subset$specimenID)]
rosmap_ind_df <- rosmap_ind_df[rosmap_ind_df$individualID %in% rosmap_meth_df$individualID, ]

#Add tissue info to methylation dataset
rosmap_meth_df$tissue <- rosmap_bio_df_subset$tissue[match(rosmap_meth_df$specimenID, rosmap_bio_df_subset$specimenID)]

#Add tissue info to individual metadata
rosmap_ind_df$tissue <- rosmap_bio_df_subset$tissue[match(rosmap_ind_df$individualID, rosmap_bio_df_subset$individualID)]

# Check if green and red .idat files present for all samples
# (idat files live in per-barcode subfolders, so list.files must recurse)
rosmap_grn_file_names <- list.files(file.path(rosmap_dir, 'raw_data'), pattern = '_Grn\\.idat$', recursive = TRUE)
rosmap_red_file_names <- list.files(file.path(rosmap_dir, 'raw_data'), pattern = '_Red\\.idat$', recursive = TRUE)

rosmap_grn_sample_names <- sub('_Grn\\.idat$', '', basename(rosmap_grn_file_names))
rosmap_red_sample_names <- sub('_Red\\.idat$', '', basename(rosmap_red_file_names))

rosmap_sample_names <- paste0(rosmap_meth_df$Sentrix_ID, '_', rosmap_meth_df$Sentrix_Row_Column)

length(intersect(rosmap_sample_names, rosmap_grn_sample_names))
length(intersect(rosmap_sample_names, rosmap_red_sample_names))

length(unique(rosmap_meth_df$Sentrix_ID))

# Samples with metadata but missing a Grn and/or Red IDAT file
rosmap_missing_idat <- union(
  setdiff(rosmap_sample_names, rosmap_grn_sample_names),
  setdiff(rosmap_sample_names, rosmap_red_sample_names)
)

# IDAT files present on disk with no matching metadata row
rosmap_missing_metadata <- union(
  setdiff(rosmap_grn_sample_names, rosmap_sample_names),
  setdiff(rosmap_red_sample_names, rosmap_sample_names)
)

cat(length(rosmap_missing_idat), "sample(s) in metadata with no matching IDAT file(s):\n")
print(rosmap_missing_idat)

cat(length(rosmap_missing_metadata), "IDAT sample(s) on disk with no matching metadata row:\n")
print(rosmap_missing_metadata)

rosmap_idat_sample_names <- union(rosmap_grn_sample_names, rosmap_red_sample_names)
plot_overlap_venn(rosmap_sample_names, rosmap_idat_sample_names, "ROSMAP")

# Remove samples missing IDAT files before writing out and summarizing
rosmap_meth_df <- rosmap_meth_df[!rosmap_sample_names %in% rosmap_missing_idat, ]
rosmap_ind_df <- rosmap_ind_df[rosmap_ind_df$individualID %in% rosmap_meth_df$individualID, ]

write.table(rosmap_ind_df, file = file.path(rosmap_out_dir, 'ROSMAP_individual_metadata_processed.csv'), sep = ',', col.names = T, row.names = F)
write.table(rosmap_meth_df, file = file.path(rosmap_out_dir, 'ROSMAP_assay_methylationArray_metadata_processed.csv'), sep = ',', col.names = T, row.names = F)

# Manifest: specimenID/individualID <-> green/red IDAT file names
rosmap_manifest <- build_idat_manifest(rosmap_meth_df, "Sentrix_ID", "Sentrix_Row_Column",
                                        rosmap_grn_file_names, rosmap_red_file_names)
write.table(rosmap_manifest, file = file.path(rosmap_out_dir, 'ROSMAP_idat_manifest.csv'), sep = ',', col.names = T, row.names = F)
```
`r length(intersect(rosmap_sample_names, rosmap_grn_sample_names))` of `r length(rosmap_sample_names)` Greenn IDAT files and `r length(intersect(rosmap_sample_names, rosmap_red_sample_names))` of `r length(rosmap_sample_names)` Red IDAT files present for ROSMAP cohort.

`r length(rosmap_missing_idat)` sample(s) in the metadata have no matching IDAT file`r if (length(rosmap_missing_idat)) paste0(": ", paste(rosmap_missing_idat, collapse = ", ")) else ""`.

`r length(rosmap_missing_metadata)` IDAT sample(s) on disk have no matching metadata row`r if (length(rosmap_missing_metadata)) paste0(": ", paste(rosmap_missing_metadata, collapse = ", ")) else ""`.

7 specimen IDs do not have associated individualIDs: PT-M5QO, TBI-AUTO73128-PT-2ZXU, PT-M5QS, TBI-AUTO73300-PT-317T, and TBI-AUTO72965-PT-318P, PT-BZBS, PT-BZRQ, PT-BZBH


## Demographics

```{r ROSMAP-demographics, echo = FALSE, results = "asis"}
demo_vars <- c("sex","race","isHispanic","apoe4Status","Braak","ageDeath")
demo_vars <- demo_vars[demo_vars %in% names(rosmap_ind_df)]
rosmap_meth_df[demo_vars] <- rosmap_ind_df[match(rosmap_meth_df$individualID, rosmap_ind_df$individualID), demo_vars]

summarize_demographics(rosmap_meth_df, "ROSMAP")
```

# ROSMAP APOE4
| Synapse ID | File name | 
|---|---|---|
| syn26607959 | RR_APOE4_datMeth.csv | 
| syn23633756 | DNA methylation Array raw data | 
| syn26607956 | RR_APOE4_metadata_assay_methylationArray.csv | 
| syn26607954 | RR_APOE4_metadata_biospecimen.csv | 
| syn73713768 | ROSMAP_clinical_harmonized.csv | 


Clinical/Individual data shared with ROSMAP (same Synapse ID as syn3191087); no separate individual metadata file for this cohort 

## Raw Data availability
```{r ROSMAP-APOE4, echo = FALSE, results = 'hide'}
apoe4_dir <- '/home/ec2-user/data/methyl_harmonization/methyl_data/ROSMAP_APOE4'
apoe4_out_dir <- '/home/ec2-user/data/methyl_harmonization/ROSMAP_APOE4/metadata/'
dir.create(apoe4_out_dir, recursive = T)
apoe4_meth_df <- read.csv(file.path(apoe4_dir, 'RR_APOE4_metadata_assay_methylationArray.csv'))
apoe4_bio_df <- read.csv(file.path(apoe4_dir, 'RR_APOE4_metadata_biospecimen.csv'))
# Individual-level clinical data is shared with ROSMAP (same harmonized file)
apoe4_ind_df <- read.csv(file.path(rosmap_dir, 'metadata/ROSMAP_clinical_harmonized.csv'))

#map the specimenID to individualID and add covariate information
apoe4_bio_df_subset <- apoe4_bio_df[apoe4_bio_df$specimenID %in% apoe4_meth_df$specimenID,]
apoe4_meth_df$individualID <- apoe4_bio_df_subset$individualID[match(apoe4_meth_df$specimenID, apoe4_bio_df_subset$specimenID)]
apoe4_ind_df <- apoe4_ind_df[apoe4_ind_df$individualID %in% apoe4_meth_df$individualID, ]

#Add tissue info to methylation dataset
apoe4_meth_df$tissue <- apoe4_bio_df_subset$tissue[match(apoe4_meth_df$specimenID, apoe4_bio_df_subset$specimenID)]

#Add tissue info to individual metadata
apoe4_ind_df$tissue <- apoe4_bio_df_subset$tissue[match(apoe4_ind_df$individualID, apoe4_bio_df_subset$individualID)]

# Validate harmonization
validate_harmonized(apoe4_ind_df, "ROSMAP APOE4")

# Check if green and red .idat files present for all samples
apoe4_grn_file_names <- list.files(file.path(apoe4_dir, 'raw_data'), pattern = '_Grn\\.idat$', recursive = TRUE)
apoe4_red_file_names <- list.files(file.path(apoe4_dir, 'raw_data'), pattern = '_Red\\.idat$', recursive = TRUE)

apoe4_grn_sample_names <- sub('_Grn\\.idat$', '', basename(apoe4_grn_file_names))
apoe4_red_sample_names <- sub('_Red\\.idat$', '', basename(apoe4_red_file_names))

apoe4_sample_names <- paste0(apoe4_meth_df$SentrixID, '_', apoe4_meth_df$SentrixRowColumn)

length(intersect(apoe4_sample_names, apoe4_grn_sample_names))
length(intersect(apoe4_sample_names, apoe4_red_sample_names))

length(unique(apoe4_ind_df$individualID))

# Samples with metadata but missing a Grn and/or Red IDAT file
apoe4_missing_idat <- union(
  setdiff(apoe4_sample_names, apoe4_grn_sample_names),
  setdiff(apoe4_sample_names, apoe4_red_sample_names)
)

# IDAT files present on disk with no matching metadata row
apoe4_missing_metadata <- union(
  setdiff(apoe4_grn_sample_names, apoe4_sample_names),
  setdiff(apoe4_red_sample_names, apoe4_sample_names)
)

cat(length(apoe4_missing_idat), "sample(s) in metadata with no matching IDAT file(s):\n")
print(apoe4_missing_idat)

cat(length(apoe4_missing_metadata), "IDAT sample(s) on disk with no matching metadata row:\n")
print(apoe4_missing_metadata)

apoe4_idat_sample_names <- union(apoe4_grn_sample_names, apoe4_red_sample_names)
plot_overlap_venn(apoe4_sample_names, apoe4_idat_sample_names, "ROSMAP APOE4")

# Remove samples missing IDAT files before writing out and summarizing
apoe4_meth_df <- apoe4_meth_df[!apoe4_sample_names %in% apoe4_missing_idat, ]
apoe4_ind_df <- apoe4_ind_df[apoe4_ind_df$individualID %in% apoe4_meth_df$individualID, ]

write.table(apoe4_ind_df, file = file.path(apoe4_out_dir, 'ROSMAP_APOE4_individual_metadata_processed.csv'), sep = ',', col.names = T, row.names = F)
write.table(apoe4_meth_df, file = file.path(apoe4_out_dir, 'ROSMAP_APOE4_assay_methylationArray_metadata_processed.csv'), sep = ',', col.names = T, row.names = F)

# Manifest: specimenID/individualID <-> green/red IDAT file names
apoe4_manifest <- build_idat_manifest(apoe4_meth_df, "SentrixID", "SentrixRowColumn",
                                       apoe4_grn_file_names, apoe4_red_file_names)
write.table(apoe4_manifest, file = file.path(apoe4_out_dir, 'ROSMAP_APOE4_idat_manifest.csv'), sep = ',', col.names = T, row.names = F)
```
`r length(intersect(apoe4_sample_names, apoe4_grn_sample_names))` of `r length(apoe4_sample_names)` Green IDAT files and `r length(intersect(apoe4_sample_names, apoe4_red_sample_names))` of `r length(apoe4_sample_names)` Red IDAT files present for ROSMAP APOE4 cohort.

`r length(apoe4_missing_idat)` sample(s) in the metadata have no matching IDAT file`r if (length(apoe4_missing_idat)) paste0(": ", paste(apoe4_missing_idat, collapse = ", ")) else ""`.

`r length(apoe4_missing_metadata)` IDAT sample(s) on disk have no matching metadata row`r if (length(apoe4_missing_metadata)) paste0(": ", paste(apoe4_missing_metadata, collapse = ", ")) else ""`.

## Demographics

```{r ROSMAP-APOE4-demographics, echo = FALSE, results = "asis"}
demo_vars <- c("sex","race","isHispanic","apoe4Status","Braak","ageDeath")
demo_vars <- demo_vars[demo_vars %in% names(apoe4_ind_df)]
apoe4_meth_df[demo_vars] <- apoe4_ind_df[match(apoe4_meth_df$individualID, apoe4_ind_df$individualID), demo_vars]

summarize_demographics(apoe4_meth_df, "ROSMAP APOE4")
```

# MOA-PAD
| Synapse ID | File name |
|---|---|
| syn34166276 | DNA methylation Array raw data |
| syn50087901 | MOA-PAD_assay_methylationArray_metadata.csv |
| syn50087874 | MOA-PAD_biospecimen_metadata.csv |
| syn50087850 | MOA-PAD_individual_metadata.csv |

## Data availability

```{r MOA-PAD, echo = FALSE, results = 'hide'}
moa_pad_dir <- '/home/ec2-user/data/methyl_harmonization/methyl_data/MOA-PAD'
moa_pad_out_dir <- '/home/ec2-user/data/methyl_harmonization/MOA-PAD/metadata/'
dir.create(moa_pad_out_dir, recursive = T)
moa_pad_meth_df <- read.csv(file.path(moa_pad_dir,'metadata/MOA-PAD_assay_methylationArray_metadata.csv'))
moa_pad_bio_df <- read.csv(file.path(moa_pad_dir,'metadata/MOA-PAD_biospecimen_metadata.csv'))
length(intersect(moa_pad_meth_df$specimenID, moa_pad_bio_df$specimenID)) #All 748 samples accounted for
moa_pad_ind_df <- read.csv(file.path(moa_pad_dir,'metadata/MOA-PAD_individual_metadata.csv'))
moa_pad_bio_df[moa_pad_bio_df$specimenID %in% moa_pad_meth_df$specimenID & moa_pad_bio_df$exclude == TRUE,] #All samples have NA as exlude 
#map the specimenID to individualID and add covariate information 
moa_pad_bio_df_subset <- moa_pad_bio_df[moa_pad_bio_df$specimenID %in% moa_pad_meth_df$specimenID,]
moa_pad_meth_df$individualID <- moa_pad_bio_df_subset$individualID[match(moa_pad_meth_df$specimenID, moa_pad_bio_df_subset$specimenID)]

#Add tissue info to methylation dataset
moa_pad_meth_df$tissue <- moa_pad_bio_df_subset$tissue[match(moa_pad_meth_df$specimenID, moa_pad_bio_df_subset$specimenID)]

#Add tissue info to individual metadata
moa_pad_ind_df$tissue <- moa_pad_bio_df_subset$tissue[match(moa_pad_ind_df$individualID, moa_pad_bio_df_subset$individualID)]

# Harmonize MOA-PAD individual metadata to the ADKP harmonized data dictionary
# In this individual file only ID/sex/race/ageDeath/diagnosis carry data;
# APOE/PMI/education/ethnicity/neuropath are NA (study reports they were assessed,
# so a fuller phenotype file likely exists on Synapse). Rows for those columns are
# kept so they populate automatically once real values are supplied.
mapping_moa_pad <- tribble(
  ~target,                 ~source,          ~type,     ~map,        ~const,       ~fn,      ~default,
  "individualID",          "individualID",   "id",      NA,          NA,           NA,       NA,
  "dataContributionGroup", NA,               "const",   NA,          "PITT-ADRC",  NA,       NA,
  "cohort",                NA,               "const",   NA,          "PITT-ADRC",  NA,       NA,
  "species",               "species",        "string",  NA,          NA,           NA,       "Human",
  "sex",                   "sex",            "map",     "sex",       NA,           NA,       NA,
  "race",                  "race",           "map",     "race",      NA,           NA,       NA,
  "isHispanic",            "ethnicity",      "map",     "hispanic",  NA,           NA,       NA,
  "ageDeath",              "ageDeath",       "age",     NA,          NA,           NA,       NA,
  "PMI",                   "pmi",            "numeric", NA,          NA,           NA,       NA,
  "apoeGenotype",          "apoeGenotype",   "apoe",    NA,          NA,           NA,       NA,
  "apoe4Status",           "apoeGenotype",   "derive",  NA,          NA,           "apoe4",  NA,
  "amyCerad",              "CERAD",          "map",     "cerad",     NA,           NA,       NA,
  "amyAny",                "amyCerad",       "derive",  NA,          NA,           "amyAny", NA,
  "amyThal",               "amyThal",        "map",     "thal",      NA,           NA,       NA,
  "amyA",                  "amyThal",        "derive",  NA,          NA,           "amyA",   NA,
  "Braak",                 "Braak",          "map",     "braak",     NA,           NA,       NA,
  "bScore",                "Braak",          "derive",  NA,          NA,           "bScore", NA
)
keep_moa_pad <- c("yearsEducation","causeDeath","mannerDeath","pH","brainWeight",
                  "diagnosis","diagnosisCriteria","tissue")

moa_pad_h <- harmonize_adkp(moa_pad_ind_df, mapping_moa_pad, keep_extra = keep_moa_pad)

#subset to individualIDs with methyltion data
moa_pad_h <- moa_pad_h[moa_pad_h$individualID %in% moa_pad_meth_df$individualID, ]

# Check if green and red .idat files present for all samples
moa_pad_grn_file_names <- list.files(file.path(moa_pad_dir, 'raw_data'), pattern = '_Grn\\.idat$', recursive = TRUE)
moa_pad_red_file_names <- list.files(file.path(moa_pad_dir, 'raw_data'), pattern = '_Red\\.idat$', recursive = TRUE)

moa_pad_grn_sample_names <- sub('_Grn\\.idat$', '', basename(moa_pad_grn_file_names))
moa_pad_red_sample_names <- sub('_Red\\.idat$', '', basename(moa_pad_red_file_names))

moa_pad_sample_names <- paste0(moa_pad_meth_df$SentrixID, '_', moa_pad_meth_df$SentrixRowColumn)

length(intersect(moa_pad_sample_names, moa_pad_grn_sample_names))
length(intersect(moa_pad_sample_names, moa_pad_red_sample_names))

length(unique(moa_pad_meth_df$individualID))

# Samples with metadata but missing a Grn and/or Red IDAT file
moa_pad_missing_idat <- union(
  setdiff(moa_pad_sample_names, moa_pad_grn_sample_names),
  setdiff(moa_pad_sample_names, moa_pad_red_sample_names)
)

# IDAT files present on disk with no matching metadata row
moa_pad_missing_metadata <- union(
  setdiff(moa_pad_grn_sample_names, moa_pad_sample_names),
  setdiff(moa_pad_red_sample_names, moa_pad_sample_names)
)

cat(length(moa_pad_missing_idat), "sample(s) in metadata with no matching IDAT file(s):\n")
print(moa_pad_missing_idat)

cat(length(moa_pad_missing_metadata), "IDAT sample(s) on disk with no matching metadata row:\n")
print(moa_pad_missing_metadata)

moa_pad_idat_sample_names <- union(moa_pad_grn_sample_names, moa_pad_red_sample_names)
plot_overlap_venn(moa_pad_sample_names, moa_pad_idat_sample_names, "MOA-PAD")

# Remove samples missing IDAT files before writing out and summarizing
moa_pad_meth_df <- moa_pad_meth_df[!moa_pad_sample_names %in% moa_pad_missing_idat, ]
moa_pad_h <- moa_pad_h[moa_pad_h$individualID %in% moa_pad_meth_df$individualID, ]

write.table(moa_pad_h, file = file.path(moa_pad_out_dir, 'MOA-PAD_individual_metadata_processed.csv'), sep = ',', row.names = F, col.names = T)
write.table(moa_pad_meth_df, file = file.path(moa_pad_out_dir, 'MOA-PAD_assay_methylationArray_metadata_processed.csv'), sep = ',', row.names = F, col.names = T)

# Manifest: specimenID/individualID <-> green/red IDAT file names
moa_pad_manifest <- build_idat_manifest(moa_pad_meth_df, "SentrixID", "SentrixRowColumn",
                                         moa_pad_grn_file_names, moa_pad_red_file_names)
write.table(moa_pad_manifest, file = file.path(moa_pad_out_dir, 'MOA-PAD_idat_manifest.csv'), sep = ',', row.names = F, col.names = T)
```
`r length(intersect(moa_pad_sample_names, moa_pad_grn_sample_names))` of `r length(moa_pad_sample_names)` Grn IDAT files and `r length(intersect(moa_pad_sample_names, moa_pad_red_sample_names))` of `r length(moa_pad_sample_names)` Red IDAT files present for MOA-PAD cohort.

`r length(moa_pad_missing_idat)` sample(s) in the metadata have no matching IDAT file`r if (length(moa_pad_missing_idat)) paste0( ": ", paste(moa_pad_missing_idat, collapse = ", ")) else ""`.

`r length(moa_pad_missing_metadata)` IDAT sample(s) on disk have no matching metadata row: `r if (length(moa_pad_missing_metadata)) paste0(": ", paste(moa_pad_missing_metadata, collapse = ", ")) else ""`.

## Demographics

```{r MOA-PAD-demographics, echo = FALSE, results = "asis"}
demo_vars <- c("sex","race","isHispanic","apoe4Status","Braak","ageDeath")
demo_vars <- demo_vars[demo_vars %in% names(moa_pad_h)]
moa_pad_meth_df[demo_vars] <- moa_pad_h[match(moa_pad_meth_df$individualID, moa_pad_h$individualID), demo_vars]

summarize_demographics(moa_pad_meth_df, "MOA-PAD")

# ---- example: harmonize and combine ----------------------------------------
# moa_pad_h <- harmonize_moa_pad(moa_pad_ind_df)   # dataContributionGroup & cohort = "Pitt"
# validate_harmonized(moa_pad_h)
#
# # union columns, then stack all three studies:
# all_cols <- union(union(names(rosmap_ind_df), names(msbb_ind_df)), names(moa_pad_h))
# fill_cols <- function(d) { for (c in setdiff(all_cols, names(d))) d[[c]] <- NA; d[all_cols] }
# combined <- dplyr::bind_rows(fill_cols(rosmap_ind_df),
#                              fill_cols(msbb_ind_df),
#                              fill_cols(moa_pad_h))
```

