---
title: "Sample-Level Methylation QC"
subtitle: "Detection p-values, intensity, predicted sex, control probes"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
mainfont: "DejaVu Sans"
fontsize: 10pt
header-includes:
  - \usepackage{titlesec}
  - \usepackage{xcolor}
  - \definecolor{darkblue}{HTML}{00008B}
  - \titleformat{\section}{\color{darkblue}\Large\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\color{darkblue}\large\bfseries}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\color{darkblue}\normalsize\bfseries}{\thesubsubsection}{1em}{}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 4,
  fig.align  = "center"
)
```
```{r libraries, include=FALSE}
library(minfi)
library(knitr)
source("/home/ec2-user/AMP-AD_methylation_harmonization/R/sample_qc.R")

# Attach reported sex onto pData(rg) from the cohort's processed individual
# metadata (not carried by the idat manifest used to compile the
# RGChannelSet), matched on individualID.
attach_sex <- function(rg, individual_csv) {
  ind <- read.csv(individual_csv, stringsAsFactors = FALSE)
  pData(rg)$sex <- ind$sex[match(pData(rg)$individualID, ind$individualID)]
  rg
}
```

Each cohort's RGChannelSet was compiled by `compile_idat_files.R`
(`<cohort>/combined_idat/<cohort>.rds`). Reported sex is merged in from
`<cohort>/metadata/<cohort>_individual_metadata_processed.csv` before QC.
`sample_qc()` flags samples on detection p-value, overall intensity, and
predicted-vs-reported sex; writes the control-probe QC report and diagnostic
plots to `<cohort>/qc/`; and returns the RGChannelSet with failing samples
removed.

# MSBB

```{r MSBB, results = 'hide'}
rg_msbb <- readRDS("/home/ec2-user/data/methyl_harmonization/MSBB/combined_idat/MSBB.rds")
rg_msbb <- attach_sex(rg_msbb,
             "/home/ec2-user/data/methyl_harmonization/MSBB/metadata/MSBB_individual_metadata_processed.csv")

dir.create("/home/ec2-user/data/methyl_harmonization/MSBB/qc",
           recursive = TRUE, showWarnings = FALSE)

res_msbb <- sample_qc(rg_msbb, sex_col = "sex",
                      out_prefix = "/home/ec2-user/data/methyl_harmonization/MSBB/qc/MSBB")

saveRDS(res_msbb$rg_clean,
        "/home/ec2-user/data/methyl_harmonization/MSBB/qc/MSBB_qc.rds")
```
```{r MSBB-summary, echo = FALSE}
kable(subset(res_msbb$qc, drop), row.names = FALSE,
      caption = "MSBB samples flagged to drop")
```

# ROSMAP

```{r ROSMAP, results = 'hide'}
rg_rosmap <- readRDS("/home/ec2-user/data/methyl_harmonization/ROSMAP/combined_idat/ROSMAP.rds")
rg_rosmap <- attach_sex(rg_rosmap,
               "/home/ec2-user/data/methyl_harmonization/ROSMAP/metadata/ROSMAP_individual_metadata_processed.csv")

dir.create("/home/ec2-user/data/methyl_harmonization/ROSMAP/qc",
           recursive = TRUE, showWarnings = FALSE)

res_rosmap <- sample_qc(rg_rosmap, sex_col = "sex",
                        out_prefix = "/home/ec2-user/data/methyl_harmonization/ROSMAP/qc/ROSMAP")

saveRDS(res_rosmap$rg_clean,
        "/home/ec2-user/data/methyl_harmonization/ROSMAP/qc/ROSMAP_qc.rds")
```
```{r ROSMAP-summary, echo = FALSE}
kable(subset(res_rosmap$qc, drop), row.names = FALSE,
      caption = "ROSMAP samples flagged to drop")
```

# ROSMAP APOE4

```{r ROSMAP-APOE4, results = 'hide'}
rg_rosmap_apoe4 <- readRDS("/home/ec2-user/data/methyl_harmonization/ROSMAP_APOE4/combined_idat/ROSMAP_APOE4.rds")
rg_rosmap_apoe4 <- attach_sex(rg_rosmap_apoe4,
                     "/home/ec2-user/data/methyl_harmonization/ROSMAP_APOE4/metadata/ROSMAP_APOE4_individual_metadata_processed.csv")

dir.create("/home/ec2-user/data/methyl_harmonization/ROSMAP_APOE4/qc",
           recursive = TRUE, showWarnings = FALSE)

res_rosmap_apoe4 <- sample_qc(rg_rosmap_apoe4, sex_col = "sex",
                              out_prefix = "/home/ec2-user/data/methyl_harmonization/ROSMAP_APOE4/qc/ROSMAP_APOE4")

saveRDS(res_rosmap_apoe4$rg_clean,
        "/home/ec2-user/data/methyl_harmonization/ROSMAP_APOE4/qc/ROSMAP_APOE4_qc.rds")
```
```{r ROSMAP-APOE4-summary, echo = FALSE}
kable(subset(res_rosmap_apoe4$qc, drop), row.names = FALSE,
      caption = "ROSMAP APOE4 samples flagged to drop")
```

# MOA-PAD

```{r MOA-PAD, results = 'hide'}
rg_moa_pad <- readRDS("/home/ec2-user/data/methyl_harmonization/MOA-PAD/combined_idat/MOA-PAD.rds")
rg_moa_pad <- attach_sex(rg_moa_pad,
                "/home/ec2-user/data/methyl_harmonization/MOA-PAD/metadata/MOA-PAD_individual_metadata_processed.csv")

dir.create("/home/ec2-user/data/methyl_harmonization/MOA-PAD/qc",
           recursive = TRUE, showWarnings = FALSE)

res_moa_pad <- sample_qc(rg_moa_pad, sex_col = "sex",
                         out_prefix = "/home/ec2-user/data/methyl_harmonization/MOA-PAD/qc/MOA-PAD")

saveRDS(res_moa_pad$rg_clean,
        "/home/ec2-user/data/methyl_harmonization/MOA-PAD/qc/MOA-PAD_qc.rds")
```
```{r MOA-PAD-summary, echo = FALSE}
kable(subset(res_moa_pad$qc, drop), row.names = FALSE,
      caption = "MOA-PAD samples flagged to drop")
```
